A Memory-Bounded, Deterministic and Terminating Semantics for the Synchronous Programming Language~\CEU

### 2018-02-24 03-semantics.tex
Corrije e Comprime prova do lema da terminação em 3.3.2

\begin{lemma}
  \label{lem.x.props-nst-i}\strut\\
  %%
  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$, for any~$p_2$:
  \begin{enumerate}[(a)]
  \item\label{lem.x.props-nst-i.a}
    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.b}
    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.c}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.d}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
  \end{enumerate}
  \medskip
  If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
  that~$\isblocked(p_1,n)$:
  \begin{enumerate}[(a)]
    \setcounter{enumi}{4}
  \item\label{lem.x.props-nst-i.e}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.f}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
  \end{enumerate}
\end{lemma}
%%
% \begin{proof}[Proof of~(a)]
%   By induction on~$i$.  The lemma is trivially true for~$i=0$,
%   as~$p_1=p_1'$, and follows directly from~\R{seq-adv} for~$i=1$.  Suppose
%   \begin{equation}
%     \label{lem.x.props-nst-i.a.eq1}
%     \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[i-1]\<p_1',n,e'>\,,
%   \end{equation}
%   for some~$i>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
%   $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
%   is false.  By~\eqref{lem.x.props-nst-i.a.eq1} and by~\R{seq-adv},
%   \begin{equation}
%     \label{lem.x.props-nst-i.a.eq2}
%     \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
%   \end{equation}
%   From~\eqref{lem.x.props-nst-i.a.eq1}, by the induction hypothesis,
%   \begin{equation}
%     \label{lem.x.props-nst-i.a.eq3}
%     \<\ceu{p_1'';\,p_2},n,e''>\nst[i-1]\<\ceu{p_1';\,p_2},n,e'>\,.
%   \end{equation}
%   From~\eqref{lem.x.props-nst-i.a.eq2} and~\eqref{lem.x.props-nst-i.a.eq3},
%   \[
%     \<\ceu{p_1;\,p_2},n,e>\nst[i]\<\ceu{p_1';\,p_2},n,e'>\,.
%   \]
%   The proofs of the other cases are similar.
% \end{proof}
%%

@@

\begin{assumption}[Syntactic restrictions]\strut
  \label{ass.x.syn-rest}
  %%
  \begin{enumerate}[(a)]
  \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contains
    no occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
    \[
      \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
    \]
    %%
  \item\label{ass.x.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all
    execution paths of~$p_1$ contain a matching~$\ceu{\Break}$ or
    an~$\ceu{\AwaitExt}$.  Consequently, for all~$n$, there are~$p_1'$
    and~$e$ such that
    \[
      \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
    \]
    where $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
  \end{enumerate}
\end{assumption}

@@

The main result of this section, Theorem~\ref{?}, is similar to
Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
general.  Before stating and proving it, we need to characterize irreducible
descriptions in general, which depend on the notions of potency of a program
and rank of a description.

@@

\begin{lemma}
  \label{lem.x.rank-nst}
  %%
  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}
\begin{proof}
  By induction on the structure of~$\nst$ derivations.  Let
  \[
    \delta=\<p,n,e>
    \quad\text{and}\quad
    \delta'=\<p',n',e'>
  \]
  be descriptions with~$\rank(\delta)=\<i,j>$ and~$\rank(\delta')=\<i',j'>$
  such that~$\delta\nst\delta'$.  Then there is a derivation~$\pi$ whose
  conclusion is~$\delta\nst\delta'$, i.e.,
  \[
    \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
  \]
  By definition of~$\nst$, $e=\nil$ and~$n=n'$.  Depending on the structure
  of program~$p$, there are~11 distinct cases to be considered.  In each one
  of them, we need to show that~$\rank(\delta)\ge\rank(\delta')$.

  We proceed to prove the case where~$p=\ceu{{p_1}\AtLoop{p_2}}$, which
  itself consists of three subcases.
  \begin{case}
    $p_1=\ceu{\Nop}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-nop}.  i.e., its conclusion is
    obtained by an application of this rule.  Hence~$p'=p_2$ and~$e'=\nil$.
    Thus
    \begin{align*}
        \rank(\delta)&=\<\pot'(p_1)+\pot'(p_2),n>\\
                     &\ge\<pot'(p_2),n>=\rank(\delta')\,.
    \end{align*}
  \end{case}
  \begin{case}
    $p_1=\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-brk}.
    Hence~$p'=p_1$ and~$e'=\nil$.  Thus
    \[
      \rank(\delta)=\rank(\delta')=\<0,n>\,.
    \]
  \end{case}
  \begin{case}
    $p_1\ne\ceu{\Nop},\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
    derivation~$\pi'$ such that
    \begin{equation}
      \label{lem.x.rank-nst-eq1}
      \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
    \end{equation}
    for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
    and~$e'=e_1'$.  There are two subcases.
    \begin{subcase}
      \label{lem.x.rank-nst-case1}
      $\pot'(p)=\pot'(p_1)$.
      %%
      Then, by Definition~\ref{def.pot}, every execution path of~$p_1$
      contains a (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
      nor can it consume an~$\ceu{\AwaitExt}$, which means that all
      execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
      or~$\ceu{\AwaitExt}$.  Hence~$\pot'(p')=\pot'(p_1')$.

      There are two further subcases depending on event~$e'$.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then $\rank(\delta)=\<pot'(p_1),n>$ and
        $\rank(\delta')=\<pot'(p_1'),n>$.  From~\eqref{lem.x.rank-nst-eq1},
        by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
        \[
          \rank(\delta)\ge\rank(\delta')\,.
        \]
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ contains one application of~\R{emit-int} which consumes
        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and which implies
        $\pot'(p_1)>\pot'(p_1')$.  Thus
        \begin{align*}
          \rank(\delta)&=\<\pot'(p_1),n>\\
                       &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
        \end{align*}
      \end{subsubcase}
    \end{subcase}
    \begin{subsubcase}
      $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
      %%
      Then some execution path in~$p_1$ does not contain a
      (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      Since~$p_1\ne\ceu{\Nop}$, a single~$\nst$ cannot restart the loop,
      which means that~$p_1'$ still contain some execution path in which
      a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ does not occur.
      Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
      similar to that of Case~\ref{lem.x.rank-nst-case1}.\qedhere
    \end{subsubcase}
  \end{case}
\end{proof}

\subsubsection*{Memory Bounds}

>>>>

\begin{lemma}
  \label{lem.x.props-nst-i}\strut\\
  %%
  If~$\<p_1,n,e>\nst[i]\<p_1',n,e'>$, for any~$p_2$:
  \begin{enumerate}[(a)]
  \item\label{lem.x.props-nst-i.a}
    $\<\ceu{p_1;\,p_2},n,e>\nst[i]\<p_1';p_2,n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.b}
    $\<\ceu{p_1\AtLoop{p_2}},n,e>\nst[i]\<\ceu{p_1'\AtLoop{p_2}},n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.c}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1'}\AtAnd{p_2}},n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.d}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}'\AtOr{p_2}},n,e'>$.
  \end{enumerate}
  \smallskip
  If~$\<p_2,n,e>\nst[i]\<p_2',n,e'>$, for any~$p_1$ such
  that~$\isblocked(p_1,n)$:
  \begin{enumerate}[(a)]
    \setcounter{enumi}{4}
  \item\label{lem.x.props-nst-i.e}
    $\<\ceu{{p_1}\AtAnd{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtAnd{p_2'}},n,e'>$.
    %%
  \item\label{lem.x.props-nst-i.f}
    $\<\ceu{{p_1}\AtOr{p_2}},n,e>\nst[i]\<\ceu{{p_1}\AtOr{p_2'}},n,e'>$.
  \end{enumerate}
\end{lemma}
%%
% \begin{proof}[Proof of~(a)]
%   By induction on~$i$.  The lemma is trivially true for~$i=0$,
%   as~$p_1=p_1'$, and follows directly from~\R{seq-adv} for~$i=1$.  Suppose
%   \begin{equation}
%     \label{lem.x.props-nst-i.a.eq1}
%     \<p_1,n,e>\nst[1]\<p_1'',n,e''>\nst[i-1]\<p_1',n,e'>\,,
%   \end{equation}
%   for some~$i>1$.  Then~$\<p_1'',n,e''>$ is not nested-irreducible, i.e.,
%   $e=\nil$ and~$p\ne{\ceu{\Nop},\ceu{\Break}}$ and~$\isblocked(p_1'',n)$
%   is false.  By~\eqref{lem.x.props-nst-i.a.eq1} and by~\R{seq-adv},
%   \begin{equation}
%     \label{lem.x.props-nst-i.a.eq2}
%     \<\ceu{p_1;\,p_2},n,e>\nst[1]\<\ceu{p_1'';\,p_2},n,e''>\,.
%   \end{equation}
%   From~\eqref{lem.x.props-nst-i.a.eq1}, by the induction hypothesis,
%   \begin{equation}
%     \label{lem.x.props-nst-i.a.eq3}
%     \<\ceu{p_1'';\,p_2},n,e''>\nst[i-1]\<\ceu{p_1';\,p_2},n,e'>\,.
%   \end{equation}
%   From~\eqref{lem.x.props-nst-i.a.eq2} and~\eqref{lem.x.props-nst-i.a.eq3},
%   \[
%     \<\ceu{p_1;\,p_2},n,e>\nst[i]\<\ceu{p_1';\,p_2},n,e'>\,.
%   \]
%   The proofs of the other cases are similar.
% \end{proof}
%%

@@

\begin{assumption}[Syntactic restrictions]\strut
  \label{ass.x.syn-rest}
  %%
  \begin{enumerate}[(a)]
  \item\label{ass.x.syn-rest.fin} If~$p=\ceu{\Fin{p_1}}$ then~$p_1$ contains
    no occurrences of~$\ceu{\AwaitExt}$, $\ceu{\AwaitInt}$, $\ceu{\Every}$,
    or~$\ceu{\Fin}$ expressions.  Consequently, for any~$n$,
    $\<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>$.
    %%
    % \[
    %   \<\clear(p_1),n,\nil>\nst[*]\<\ceu{\Nop},n,\nil>\,.
    % \]
    %%
  \item\label{ass.x.syn-rest.loop} If~$p=\ceu{\Loop{p_1}}$ then all
    execution paths of~$p_1$ contain a matching~$\ceu{\Break}$ or
    an~$\ceu{\AwaitExt}$.  Consequently, for all~$n$, there are~$p_1'$
    and~$e$ such that
    $\<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>$,
    %%
    % \[
    %   \<\ceu{\Loop{p_1}},n,\nil>\nst[*]\<p_1',n,e>\,,
    % \]
    %%
    where $p_1'=\ceu{{\Break}\AtLoop{p_1}}$ or~$\isblocked(p_1',n)$\,.
  \end{enumerate}
\end{assumption}

@@

The main result of this section, Theorem~\ref{thm.x.term}, is similar to
Theorem~\ref{thm.x.term-nst-*} but applies to transitions~$\trans$ in
general.  Before stating and proving it, we need to characterize irreducible
descriptions in general.  This characterization depends on the notions of
potency and rank.

@@

\begin{lemma}
  \label{lem.x.rank-nst}
  %%
  If~$\delta\nst\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{lemma}
\begin{proof}
  By induction on the structure of~$\nst$ derivations.  Let
  \[
    \delta=\<p,n,e>
    \quad\text{and}\quad
    \delta'=\<p',n',e'>
  \]
  be descriptions with~$\rank(\delta)=\<i,j>$ and~$\rank(\delta')=\<i',j'>$
  such that~$\delta\nst\delta'$.  Then there is a derivation~$\pi$ whose
  conclusion is~$\delta\nst\delta'$, i.e.,
  $\pi\Vdash\<p,n,e>\nst\<p',n',e'>$.
  %%
  % \[
  %   \pi\Vdash\<p,n,e>\nst\<p',n',e'>\,.
  % \]
  %%
  By definition of~$\nst$, $e=\nil$ and~$n=n'$.  Depending on the structure
  of program~$p$, there are~11 distinct cases to be considered.  In each one
  of them, we need to show that~$\rank(\delta)\ge\rank(\delta')$.

  We proceed to prove the case where~$p=\ceu{{p_1}\AtLoop{p_2}}$, which
  itself consists of three subcases.
  \begin{case}
    $p_1=\ceu{\Nop}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-nop}.  i.e., its conclusion is
    obtained by an application of this rule.  Hence~$p'=p_2$ and~$e'=\nil$.
    Thus
    \begin{align*}
        \rank(\delta)=\<\pot'(p_1)+\pot'(p_2),n>
                     \ge\<pot'(p_2),n>=\rank(\delta')\,.
    \end{align*}
  \end{case}
  \begin{case}
    $p_1=\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-brk}.
    Hence~$p'=p_1$ and~$e'=\nil$.  Thus
    $\rank(\delta)=\rank(\delta')=\<0,n>$.
    %%
    % \[
    %   \rank(\delta)=\rank(\delta')=\<0,n>\,.
    % \]
    %%
  \end{case}
  \begin{case}
    $p_1\ne\ceu{\Nop},\ceu{\Break}$.
    %%
    Then~$\pi$ is an instance of~\R{loop-adv}.  Hence there is a
    derivation~$\pi'$ such that
    \begin{equation}
      \label{lem.x.rank-nst-eq1}
      \pi'\Vdash\<p_1,n,\nil>\nst\<p_1',n,e_1'>\,,
    \end{equation}
    for some~$p_1'$ and~$e_1'$.  Thus~$p'=\ceu{p_1'\AtLoop{\,p_2}}$
    and~$e'=e_1'$.  There are two subcases.
    \begin{subcase}
      \label{lem.x.rank-nst-case1}
      $\pot'(p)=\pot'(p_1)$.
      %%
      Then, by Definition~\ref{def.pot}, every execution path of~$p_1$
      contains a (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      A single~$\nst$ cannot terminate the loop, since~$p_1\ne\ceu{\Break}$,
      nor can it consume an~$\ceu{\AwaitExt}$, which means that all
      execution paths in~$p_1'$ still contain a~$\ceu{\Break}$
      or~$\ceu{\AwaitExt}$.  Hence~$\pot'(p')=\pot'(p_1')$.

      There are two further subcases depending on event~$e'$.
      \begin{subsubcase}
        $e'=\nil$.
        %%
        Then $\rank(\delta)=\<pot'(p_1),n>$ and
        $\rank(\delta')=\<pot'(p_1'),n>$.  From~\eqref{lem.x.rank-nst-eq1},
        by the induction hypothesis, $\pot'(p_1)\ge\pot(p_1')$.  Thus
        $\rank(\delta)\ge\rank(\delta')$.
        %%
        % \[
        %   \rank(\delta)\ge\rank(\delta')\,.
        % \]
        %%
      \end{subsubcase}
      \begin{subsubcase}
        $e'\ne\nil$.
        %%
        Then~$\pi'$ contains one application of~\R{emit-int} which consumes
        one~$\ceu{\EmitInt(e')}$ expression from~$p_1$ and which implies
        $\pot'(p_1)>\pot'(p_1')$.  Thus
        \[
          \hskip1em\rank(\delta)\!=\!\<\pot'(p_1),n>
          >\<\pot'(p_1'),n+1>\!=\!\rank(\delta')\,.
        \]
        %%
        % \begin{align*}
        %   \rank(\delta)&=\<\pot'(p_1),n>\\
        %                &>\<\pot'(p_1'),n+1>=\rank(\delta')\,.
        % \end{align*}
        %%
      \end{subsubcase}
    \end{subcase}
    \begin{subsubcase}
      $\pot'(p)=\pot'(p_1)+\pot'(p_2)$.
      %%
      Then some execution path in~$p_1$ does not contain a
      (matching)~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ expression.
      Since~$p_1\ne\ceu{\Nop}$, a single~$\nst$ cannot restart the loop,
      which means that~$p_1'$ still contain some execution path in which
      a~$\ceu{\Break}$ or~$\ceu{\AwaitExt}$ does not occur.
      Hence~$\pot'(p')=\pot'(p_1')+\pot'(p_2)$.  The rest of this proof is
      similar to that of Case~\ref{lem.x.rank-nst-case1}.\qedhere
    \end{subsubcase}
  \end{case}
\end{proof}

The next theorem is a generalization of Lemma~\ref{lem.x.rank-nst}
for~$\nst[*]$ transitions.  Its proof is by induction on~$i$.

\begin{theorem}
  \label{thm.x.rank-nst-*}
  %%
  If~$\delta\nst[*]\delta'$ then~$\rank(\delta)\ge\rank(\delta')$.
\end{theorem}

We now state and prove the main result of this section,
Theorem~\ref{thm.x.term}, viz., the termination theorem for~$\trans[*]$.
The idea of the proof is that a series of~$\nst$ or~$\out$ transitions
eventually decrease the rank of the current description until an irreducible
description is reached.  This irreducible description is the final result of
the reaction.
%%
\begin{theorem}[Termination]\label{thm.x.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$\rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.x.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.x.rank-nst-*},
    \[
      \<i,j>=\<0,0>\ge\rank(\delta')\,,
    \]
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases depending on whether or not $\delta$~is
    nested-irreducible.
    \begin{case}
      \label{thm.x.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      If~$j=0$ then, by Definition~\ref{def.x.H}, $\delta_\#$, and
      thus~$\delta\trans[0]\delta_\#$.  Suppose that is not the case, i.e.,
      that $j>0$.  Then there are two subcases.
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nonnil}
        $e\ne\nil$.
        %%
        Then, by~\R{out} and by Theorem~\ref{thm.x.term-nst-*}, there
        are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
        $\delta\outpush\delta_1'\nst[*]\delta'_\Hnst$.  Thus, by
        Lemma~\ref{lem.x.rank-out} and by Theorem~\ref{thm.x.rank-nst-*},
        \begin{align*}
          \rank(\delta)=\rank(\delta_1')=\<i,j>
                                        \ge\rank(\delta')=\<i',j'>\,.
        \end{align*}
        If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof is
        similar to that of Case~\ref{thm.x.term.Hnst-j>0-nil} below.
        Otherwise, if~$e'\ne\nil$ then $i>i'$, since an~$\ceu{\EmitInt(e')}$
        was consumed by the nested transitions.  Thus,
        $\rank(\delta)>\rank(\delta')$.  By the induction hypothesis,
        $\delta'\trans[*]\delta''_\#$, for some~$\delta''_\#$.  Therefore,
        $\delta\trans[*]\delta''_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.x.term.Hnst-j>0-nil}
        $e=\nil$.
        %%
        Then, since~$j>0$, $\delta\outpop\delta'$, for some~$\delta''$.  By
        item~\eqref{lem.rank-out-pop} of Lemma~\ref{lem.rank-out},
        $\rank(\delta)>\rank(\delta')$.  Hence, by the induction hypothesis,
        there is a~$\delta''_\#$ such that~$\delta'\trans[*]\delta''_\#$.
        Therefore, $\delta\trans[*]\delta''_\#$.
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorems~\ref{thm.x.rank-nst-*}
      and~\ref{thm.x.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.x.term.Hnst} above.\qedhere
    \end{case}
  \end{induction}
\end{proof}

\subsubsection*{Memory bounds}

### 2018-02-24 06-appendix.tex
Corrije e Comprime prova do lema da terminação em 3.3.2

\begin{theorem}[Termination]\label{thm.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  %%
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.rank-nst-*},
    \[
      \<i,j>=\<0,0>\ge\rank(\delta')\,,
    \]
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases.
    \begin{case}
      \label{thm.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      There are two cases.
      \begin{subcase}
        \label{thm.term.Hnst-j0}
        $j=0$.  By Definition~\ref{def.H}, $\delta_\#$.
        Thus~$\delta\trans[0]\delta_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.term.Hnst-j>0}
        $j>0$.
        %%
        There are two subcases.
        \begin{subsubcase}
          \label{thm.term.Hnst-j>0-nonnil}
          $e\ne\nil$.
          %%
          Then, by~\R{out} and by Theorem~\ref{thm.term-nst-*}, there
          are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
          \[
            \delta\outpush\delta_1'\nst[*]\delta'_\Hnst\,.
          \]
          Thus, by Lemma by item~\eqref{lem.rank-out-push} of
          Lemma~\ref{lem.rank-out} and by Theorem~\ref{thm.rank-nst-*},
          \begin{align*}
            \rank(\delta)=\rank(\delta_1')&=\<i,j>\\
                                          &\ge\rank(\delta')=\<i',j'>\,.
          \end{align*}
          If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof
          is similar to that of Case~\ref{thm.term.Hnst-j>0-nil}.
          Otherwise, if~$e'\ne\nil$ then $i>i'$, since
          an~$\ceu{\EmitInt(e')}$ was consumed by the nested transitions.
          Thus,
          \[
            \rank(\delta)>\rank(\delta')\,.
          \]
          By the induction hypothesis, $\delta'\trans[*]\delta''_\#$, for
          some~$\delta''_\#$.  Therefore, $\delta\trans[*]\delta''_\#$.
        \end{subsubcase}
        \begin{subsubcase}
          \label{thm.term.Hnst-j>0-nil}
          $e=\nil$.
          %%
          Then, since~$j>0$, $\delta\outpop\delta'$, for
          some~$\delta''$.  By item~\eqref{lem.rank-out-pop} of
          Lemma~\ref{lem.rank-out},
          \[
            \rank(\delta)>\rank(\delta')\,.
          \]
          Hence, by the induction hypothesis, there is a~$\delta''_\#$ such
          that~$\delta'\trans[*]\delta''_\#$.
          Therefore, $\delta\trans[*]\delta''_\#$.
        \end{subsubcase}
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      There are two subcases.
      \begin{subcase}
        $e\ne\nil$.  Similar to Case~\ref{thm.term.Hnst-j>0-nonnil}.
      \end{subcase}
      \begin{subcase}
        $e=\nil$.  Then, by Theorems~\ref{thm.rank-nst-*}
        and~\ref{thm.rank-nst-*} there is a~$\delta'_\Hnst$ such
        that~$\delta\nst[*]\delta'_\Hnst$
        with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
        is similar to that of Case~\ref{thm.term.Hnst}.\qedhere
      \end{subcase}
    \end{case}
  \end{induction}
\end{proof}

>>>>

\begin{theorem}[Termination]\label{thm.term}\strut\\
  %%
  For any~$\delta$, there is a~$\delta'_\#$ such
  that~$\delta\trans[*]\delta'_\#$.
\end{theorem}
\begin{proof}
  By lexicographic induction on~$rank(\delta)$.  Let~$\delta=\<p,n,e>$ and
  $\rank(\delta)=\<i,j>$.
  %%
  \begin{basis}
    If~$\<i,j>=\<0,0>$ then~$\delta$ cannot be advanced by~$\out$, as~$j=0$
    implies~$e=\nil$ and~$n=0$ (neither~\R{push} nor~\R{pop} can be
    applied).  There are two possibilities:  either~$\delta$ is
    nested-irreducible or it is not.  In the first case, the theorem is
    trivially true, as~$\delta\nst[0]\delta_\Hnst$.  Suppose~$\delta$ is not
    nested-irreducible.  Then, by Theorem~\ref{thm.term-nst-*},
    $\delta\nst[*]\delta'_\Hnst$ for some~$\delta'_\Hnst$.  By
    Theorem~\ref{thm.rank-nst-*},
    \[
      \<i,j>=\<0,0>\ge\rank(\delta')\,,
    \]
    which implies~$\rank(\delta')=\<0,0>$.
  \end{basis}
  \begin{induction}
    Let~$\<i,j>\ne\<0,0>$.
    %%
    There are two subcases.
    \begin{case}
      \label{thm.term.Hnst}
      $\delta$~is nested-irreducible.
      %%
      There are two cases.
      \begin{subcase}
        \label{thm.term.Hnst-j0}
        $j=0$.  By Definition~\ref{def.H}, $\delta_\#$.
        Thus~$\delta\trans[0]\delta_\#$.
      \end{subcase}
      \begin{subcase}
        \label{thm.term.Hnst-j>0}
        $j>0$.
        %%
        There are two subcases.
        \begin{subsubcase}
          \label{thm.term.Hnst-j>0-nonnil}
          $e\ne\nil$.
          %%
          Then, by~\R{out} and by Theorem~\ref{thm.term-nst-*}, there
          are~$\delta_1'$ and~$\delta'_\Hnst=\<p',n+1,e'>$ such that
          \[
            \delta\outpush\delta_1'\nst[*]\delta'_\Hnst\,.
          \]
          Thus, by item~\eqref{lem.rank-out-push} of
          Lemma~\ref{lem.rank-out} and by Theorem~\ref{thm.rank-nst-*},
          \begin{align*}
            \rank(\delta)=\rank(\delta_1')&=\<i,j>\\
                                          &\ge\rank(\delta')=\<i',j'>\,.
          \end{align*}
          If~$e'=\nil$, then $i=i'$ and~$j=j'$, and the rest of this proof
          is similar to that of Case~\ref{thm.term.Hnst-j>0-nil}.
          Otherwise, if~$e'\ne\nil$ then $i>i'$, since
          an~$\ceu{\EmitInt(e')}$ was consumed by the nested transitions.
          Thus,
          \[
            \rank(\delta)>\rank(\delta')\,.
          \]
          By the induction hypothesis, $\delta'\trans[*]\delta''_\#$, for
          some~$\delta''_\#$.  Therefore, $\delta\trans[*]\delta''_\#$.
        \end{subsubcase}
        \begin{subsubcase}
          \label{thm.term.Hnst-j>0-nil}
          $e=\nil$.
          %%
          Then, since~$j>0$, $\delta\outpop\delta'$, for
          some~$\delta''$.  By item~\eqref{lem.rank-out-pop} of
          Lemma~\ref{lem.rank-out},
          \[
            \rank(\delta)>\rank(\delta')\,.
          \]
          Hence, by the induction hypothesis, there is a~$\delta''_\#$ such
          that~$\delta'\trans[*]\delta''_\#$.
          Therefore, $\delta\trans[*]\delta''_\#$.
        \end{subsubcase}
      \end{subcase}
    \end{case}
    \begin{case}
      $\delta$~is not nested-irreducible.
      %%
      Then~$e=\nil$ and, by Theorems~\ref{thm.rank-nst-*}
      and~\ref{thm.rank-nst-*}, there is a~$\delta'_\Hnst$ such
      that~$\delta\nst[*]\delta'_\Hnst$
      with~$\rank(\delta)\ge\rank(\delta'_\Hnst)$.  The rest of this proof
      is similar to that of Case~\ref{thm.term.Hnst}.\qedhere
    \end{case}
  \end{induction}
\end{proof}
